\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-25 paper.tex

% Other representations of ARGs, beyond straightforward implementations
% of a Griffiths-style
% eARG~\citep[e.g.][]{ignatieva2021kwarg,rasmussen2014genome}
% and \texttt{tskit}'s gARG format, are of course possible.
% The ARGON simulator~\citep{palamara2016argon}, for example,
% stores ARGs
% ARG-Needle~\citep{zhan2023towards}, for example, uses a format based
% on hap

@@

\section{Discussion}
\label{sec-discussion}
Recent breakthroughs have finally made large-scale ARG inference
feasible in practice, leading to a surge of interest
in development of ARG-based inference methods, their evaluation
and application.
The prospect of ARGs being used routinely within population
and statistical genetics is tantalising,
but in reality there is substantial work to be done to
enable this.
A necessary first step is a degree of terminological clarity.
As discussed in Appendix~\ref{sec-arg-history}, the term
``ancestral recombination graph" has several
subtly different interpretations, depending on context.
The trend to decouple ARGs from their original definition
within the context of stochastic
processes and instead use the term as a more general representation of any
recombinant genetic ancestry seems useful, and we have
tried to clarify and systematise it here. Thus
we can think of an ARG as any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG (gARG) encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which is more limited in what can be described,
because of its origins in an idealised mathematical model,
and the assumption of having complete information about
recombination events.

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While this has proven to be a useful and
robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
many modern datasets have properties that grossly
violate its assumptions.
For instance, several human datasets now consist of hundreds of thousands of
genomes~\citep{turnbull2018hundred, bycroft2018genome,
karczewski2020mutational,tanjo2021practical,
halldorsson2022sequences},
and so sample size is substantially \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and
myriad interconnections of migration, were somehow estimated and
provided as input, ARGs sampled from
the coalescent cannot capture the complexities of realistic family structures, which often involve overlapping generations and extreme variation in lifetime reproductive success. Flexible ARG encoding should accommodate such cases.

Given that there is no single model that can capture the complexities
of modern datasets across species (and taxa; see discussion of SARS-CoV-2
below), we cannot rely on sampling from a distribution to capture
the uncertainty in our ARG estimates. Indeed, even if the coalescent
were a suitable model for datasets such as UKB and we had a means of
sampling from its distribution of ARGs at the scale of hundreds of thousands
of samples, we could only explore the tiniest corner of such an incomprehensibly
large space of possibilities.
Therefore, it is important to systematically
describe and utilise uncertainty about ARG inference.
One approach, enabled by the gARG encoding described here, is to allow
nodes to have (rootward) in-degree greater than two (polytomies
representing uncertainty over the ordering of coalescence events) or
out-degree greater than two (representing uncertainty over the ordering
of multiple recombination events). Development of other methods to capture, for example,
uncertainty about node ages and recombination breakpoint positions, is an important
aspect of future work.

The timing, positions, and even the number of recombination events is generally
not possible to infer precisely from genome sequencing data. For ARGs under
coalescent-based models, the proportion of recombination events that change the
ARG topology grows very slowly with sample size \citep{hein2004gene}, and of those
events only a small proportion are actually detectable from the data, assuming
human-like mutation rates \citep{myers2002detection,hayman2023recoverability}.
Even when a recombination event \emph{is} detectable, its timing and breakpoint
position can only be inferred approximately, depending on how much information
can be elucidated from mutations in the surrounding genomic region. For general
ARGs, for instance those representing pedigrees, or genealogies of organisms with
low mutation rates and complex patterns of recombination, such inference can be
even more challenging. The fact that the eARG encoding \emph{requires}
precise information about recombination is therefore a fundamental limitation, which
is a central message of this paper.
% This might be where we mention Fisher's junctions and IBD?
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/121

% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully
precise ARGs from data,
we should also consider the value that such precise estimates provide
for downstream applications.
% To date, few make use such detailed recombination information.
Most applications work by examining local trees independently.
For example, the \relate\ selection test~\citep{speidel2019method}
obtains $p$-values by computing clade size probabilities conditional
on the timing of coalescence events in a given local tree.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\cite{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
Similarly, \cite{fan2023likelihood} compute the likelihood
of an ARG under a particular demographic model as the product
over a sample of widely-separated local trees, assumed to be independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account. Thus,
while SIA does use some of the information about local tree correlation,
clearly much of the detail about recombination
events in an ARG is lost.
The \texttt{tsdate} algorithm (and related naive estimator
of ancestral location) uses much more between-tree
information~\citep{wohns2022unified}, explicitly using the gARG
encoding to reason about ancestral nodes.
The main application for fully precise ARGs thus far has been
to compute a likelihood under the
coalescent~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian,
guo2022recombination},
which currently requires the details of all recombination
events to be known.

% Note: haven't said anything about the coalescent here. Do we need to?
The advantages of a model-agnostic representation that naturally
incorporates uncertainty about the ordering of events in an ARG are well-illustrated by~\cite{zhan2023towards}, who
inferred ARGs using millions of SARS-CoV-2 sequences from the
GISAID database~\citep{shu2017gisaid}. In contrast to typical human sequencing datasets, the SARS-CoV-2 data is
sampled continuously through time, sometimes with tens of thousands of sequences collected per day, with relatively little genetic diversity to
distinguish them. The reconstructed ARGs thus contain polytomies (representing
both the uncertainty around coalescence times and the presence of true
multiple mergers), non-leaf sample nodes (sequences with descendants
also present in the dataset) and unary nodes (samples that have
one other sampled descendant). Recombination is known
to be an important factor in the evolution of SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing},
and the ARGs also contain recombination nodes.
Because parental sequences
are generally never sampled themselves, and often
a recombinant strain is the product of multiple recombination events,
uncertainty around this is captured by recording the ancestry
of each part of the recombinant sequence without arbitrarily assigning
times or orderings for these events. This demonstrates the flexibility of the gARG
encoding in capturing recombinant genealogies regardless of the underlying
(possibly unknowable) generative evolutionary process.

This view of ARGs,
decoupled from generative models and
without the hard requirement
of complete precision on all historical events, may clarify inference goals and improve
methods for evaluation.
In most cases,
ARG inference is evaluated by simulating data from a known ground truth ARG,
and comparing this to the inferred version via pairwise comparison of
local trees along the genome, for example
using tree distance
metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping},
as described by \citet{kuhner2015assessing}.
% While this provides valuable insights into the method's performance,
% it is not clear how well these tree distance metrics reflect
% performance on real data.
In comparing tree-by-tree along the genome, the effects of recombination
are only incorporated in an indirect manner through the correlations
between the local trees, rather than through directly taking into account
the persistence of nodes and edges across multiple trees.
The performance of tree distance metrics varies by
application~\citep{kuhner2015assessing}, and the correct approach
to handling subtleties such as polytomies
is an open question~\citep{kelleher2019inferring,zhang2023biobank}.
% Is this true? These guys claim RF takes linear time:
% https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-020-07011-0
Tree distance metrics usually have at least $O(n^2)$ time complexity
and therefore cannot be
applied to the very large sample sizes currently of interest.
A recent trend has been to move away from such
tree distance-based approaches and to examine more properties of the inferred ARGs.
\citet{brandt2021evaluation} focussed on the
distributions of pairwise MRCA times;
\citet{deng2021distribution} derived an approximation to the distribution of the
waiting distances between local trees;
\citet{ignatieva2023distribution}
derived the distribution of the genomic span of an edge and that of a
clade of samples. In each case, simulation studies demonstrated
substantial differences between these quantities in simulated and reconstructed ARGs,
that were not captured using tree-by-tree comparisons.
% Not sure this is the right place for this sentence
These evaluations have been based on
ground truth data from highly idealised simulations, with sample sizes limited to at most a few thousand (typically much fewer).
The effects of the richness of real data
on modern biobank-scale or multi-generational pedigree datasets
are almost entirely unknown (beyond studies using very simplistic error models).
The development of ARG evaluation metrics that take into account more of the
global topology and can be applied to large ARGs would be a
valuable and timely addition to the field.

Interest in ARG inference methods and downstream applications is
burgeoning, with exciting developments arriving at ever-increasing pace.
Without agreement on basic terminology and some standardisation
on data formats, however, the ARG revolution may falter.
For ARG-based methods to achieve mainstream status, we require
a rich supporting software ecosystem.
Ideally, this would comprise a wide range of different
inference methods specialised to different organisms, types
and scales of data and
% I'm thinking about things like Espalier specifically wanting
% to characterise the recombs, vs Relate mainly wanting to
% get good local trees
specific inference goals, sharing a common ARG interchange format.
The output of these diverse inference methods could then be
processed by numerous downstream applications,
enabling users to freely compare the outputs of different
inference methods on their data, and the result of different
analyses on these inferences.
Earlier efforts to standardise ARG interchange shared this vision,
but did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems with converting data from one program's
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
Current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package,
which is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse software ecosystem.
The gARG encoding described here is a significant generalisation
of classical concepts, capable of describing even the bewildering complexity
of contemporary datasets and describing important forms of temporal uncertainty.
The \texttt{tskit} library is a high-quality open-source gARG implementation,
with proven efficiency and
scalability~\citep[e.g.][]{anderson2022genes,zhan2023towards},
that is already in widespread use.
Adopting it as a community standard,
and utilising its rich suite of operations,
may accelerate progress and help to
establish ARG methods within mainstream genetics.

>>>>

@@

\section{Discussion}
\label{sec-discussion}
Recent breakthroughs have finally made large-scale ARG inference
feasible in practice, leading to a surge of interest
in inference methods, their evaluation, and their application to biological questions.
The prospect of ARGs being used routinely within population
and statistical genetics is tantalising,
but in reality there is substantial work to be done to
enable this.
A necessary first step is a degree of terminological clarity.
As discussed in Appendix~\ref{sec-arg-history}, the term
``ancestral recombination graph" has several
subtly different interpretations, depending on context.
The trend to decouple ARGs from their original definition
within the context of stochastic
processes and instead use the term as a more general representation of any
recombinant genetic ancestry seems useful, and we have
tried to clarify and systematise it here. Thus
we can think of an ARG as any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG (gARG) encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG) encoding, which is more limited in what can be described.
Importantly, gARGs do not require fully precise estimates of
ancestral recombination events,
and allow us to directly express important forms of temporal uncertainty.

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process (henceforth, ``coalescent'')
is an important step.
While the coalescent has proven to be a useful and
robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
many modern datasets have properties that grossly
violate its assumptions.
One key assumption is that
sample size $n$ is much less than the effective population size, $N_e$.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{turnbull2018hundred, bycroft2018genome,
karczewski2020mutational,tanjo2021practical,
halldorsson2022sequences},
and so sample size is an order of magnitude \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from this
assumption, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
A model assuming a single $N_e$ would be a
drastic over-simplification of course, but
even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and myriad
interconnections of migration, were somehow estimated and provided as input,
ARGs sampled from the coalescent cannot capture the complexities
of family structure in these
datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
Another core assumption of the coalescent model is that the genome
(or at least the region under study) is short enough that the number of extant
lineages remains much smaller than $N_e$ at all times.
High-quality whole genome assemblies are now available
for many species % Can we say something more concrete here?
and projects are under way to obtain
them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth},
and so we can expect inferred ARGs to routinely span whole chromosomes.

Recent large-scale methods have simplified the inference problem by
making a single, deterministic best-guess at ARG
inference~\citep{kelleher2019inferring,speidel2019method,zhan2023towards}.
Even under strict parsimony conditions and for small sample sizes, the
% Citation? This is something people have figured out, right?
number of plausible ARGs compatible with a given dataset is vast.
Thus, although it is clearly an oversimplification to arbitrarily
choose \emph{one} best guess, it is not clear that sampling more
when sample sizes are large will achieve much.
At the scale of millions of samples,
we could only ever explore the tiniest corner of the incomprehensibly
large space of likely ARGs.
Therefore, it is important to systematically
describe and utilise uncertainty about ARG inference, and to
incorporates uncertainty encountered during inference into the returned ARG.
One approach, enabled by the gARG encoding described here, is to allow
nodes to have (rootward) in-degree greater than two (polytomies
representing uncertainty over the ordering of coalescence events) or
out-degree greater than two (representing uncertainty over the ordering
of multiple recombination events). Development of other methods to capture, for example,
uncertainty about node ages and recombination breakpoint positions, is an important
aspect of future work. Fully utilising such uncertainty in downstream
applications is also vital, of course.

The timing, positions, and even the number of recombination events is generally
not possible to infer precisely from genome sequencing data. Under
coalescent-based models, the proportion of recombination events that change the
ARG topology grows very slowly with sample size \citep{hein2004gene}, and of those
events only a small proportion are actually detectable from the data, assuming
human-like mutation rates \citep{myers2002detection,hayman2023recoverability}.
Even when a recombination event \emph{is} detectable, its timing and breakpoint
position can only be inferred approximately, depending on how much information
can be elucidated from mutations in the surrounding genomic region.
% For general
% ARGs, for instance those representing pedigrees, or genealogies of organisms with
% low mutation rates and complex patterns of recombination, such inference can be
% even more challenging.
The fact that the eARG encoding \emph{requires}
precise information about recombination is therefore a fundamental limitation.
% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully
precise ARGs from data,
we should also consider the value that such precise estimates provide
for downstream applications.
Many applications work by examining local trees independently.
For example, the \relate\ selection test~\citep{speidel2019method}
obtains $p$-values by computing clade size probabilities conditional
on the timing of coalescence events in a given local tree.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\cite{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
Similarly, \cite{fan2023likelihood} compute the likelihood
of an ARG under a particular demographic model as the product
over a sample of widely-separated local trees, assumed to be independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account.
Thus, while SIA takes advantage of information about local tree correlation,
it is in quite an indirect way, and
clearly much of the detail about recombination events in an ARG is lost.
The \texttt{tsdate} algorithm (and related naive estimator
of ancestral location) uses much more between-tree
information~\citep{wohns2022unified}, explicitly using the gARG
encoding to reason about ancestral nodes.
The main application for fully precise ARGs thus far has been
to compute a likelihood under the
coalescent~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian,
guo2022recombination},
which currently requires the details of all recombination
events to be known.

The advantages of a model-agnostic representation that naturally
incorporates uncertainty about the ordering of events in an ARG
are well-illustrated by~\cite{zhan2023towards}, who
inferred ARGs using millions of SARS-CoV-2 sequences from the
GISAID database~\citep{shu2017gisaid}.
In contrast to typical human sequencing datasets, the SARS-CoV-2 data is
sampled continuously through time, sometimes with tens of thousands of sequences collected per day, with relatively little genetic diversity to
distinguish them. The reconstructed ARGs thus contain polytomies (representing
both the uncertainty around coalescence times and the presence of true
multiple mergers), non-leaf sample nodes (sequences with descendants
also present in the dataset) and unary nodes (samples that have
one direct descendant).
Recombination is an important factor in the evolution of
SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing},
% Tried to jazz this up a bit, was quite dry
and the inferred ARGs contain an unprecedented level of detail about
the combined processes of viral mutation and recombination.
Because parental sequences
are generally never sampled themselves, and often
a recombinant strain is the product of multiple recombination events,
uncertainty around this is captured by recording the ancestry
of each part of the recombinant sequence without arbitrarily assigning
times or orderings for these events.

This view of ARGs,
decoupled from generative models and
without the hard requirement
of complete precision on all historical events, may clarify inference goals and improve
methods for evaluation.
In most cases,
ARG inference is evaluated by simulating data from a known ground truth ARG,
and comparing this to the inferred version via pairwise comparison of
local trees along the genome, for example
using tree distance
metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping},
as described by \citet{kuhner2015assessing}.
% While this provides valuable insights into the method's performance,
% it is not clear how well these tree distance metrics reflect
% performance on real data.
In comparing tree-by-tree along the genome, the effects of recombination
are incorporated in quite an indirect manner through the correlations
between the local trees, rather than through directly taking into account
the persistence of nodes and edges across multiple trees.
The performance of tree distance metrics varies by
application~\citep{kuhner2015assessing}, and the correct approach
to handling subtleties such as polytomies
is an open question~\citep{kelleher2019inferring,zhang2023biobank}.
% Is this true? These guys claim RF takes linear time:
% https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-020-07011-0
Tree distance metrics often have at least $O(n^2)$ time complexity
and therefore cannot be
applied to the very large sample sizes currently of interest.
A recent trend has been to move away from such
tree distance-based approaches and to examine more
properties of the inferred ARGs, such as
distributions of pairwise MRCA times~\citep{brandt2021evaluation},
waiting distances between local trees~\citep{deng2021distribution},
and the genomic span of an edge or clade of
samples~\citep{ignatieva2023distribution}.
In each case, simulation studies demonstrated
substantial differences between these quantities in simulated and reconstructed ARGs,
that were not captured using tree-by-tree comparisons.
Evaluations to-date have almost all been based
on ground truth data from highly idealised simulations,
with sample sizes limited to at most a few thousand (typically much fewer).
Beyond the effects of very simplistic error models~\cite[e.g.][]{kelleher2019inferring},
the effects of the richness of real data
at biobank-scale on ARG inference are almost entirely unknown.
The development of ARG evaluation metrics that take into account more of the
global topology and can be applied to large ARGs would be a
valuable and timely addition to the field.

Interest in ARG inference methods and downstream applications is
burgeoning, with exciting developments arriving at ever-increasing pace.
Without agreement on basic terminology and some standardisation
on data formats, however, the ARG revolution may falter.
For ARG-based methods to achieve mainstream status, we require
a rich supporting software ecosystem.
Ideally, this would comprise a wide range of different
inference methods specialised to different organisms, types
and scales of data and
% I'm thinking about things like Espalier specifically wanting
% to characterise the recombs, vs Relate mainly wanting to
% get good local trees
targetting particular inference goals, sharing a common ARG interchange format.
The output of these diverse inference methods could then be
processed by numerous downstream applications,
enabling users to freely compare the outputs of different
inference methods on their data, and the result of different
analyses on these inferences.
Earlier efforts to standardise ARG interchange shared this vision,
but did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems with converting data from one program's
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
Current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package,
which is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse software ecosystem.
The gARG encoding described here is a significant generalisation
of classical concepts, capable of describing even the bewildering complexity
of contemporary datasets and capturing important forms of temporal uncertainty.
The \texttt{tskit} library is a high-quality open-source gARG implementation,
with proven efficiency and
scalability~\citep[e.g.][]{anderson2022genes,zhan2023towards},
that is already in widespread use.
Adopting it as a community standard,
and utilising its rich suite of operations,
may accelerate progress and help to
establish ARG methods within mainstream genetics.
